systemfit: A Package to Estimate\\ Simultaneous Equation Systems in \proglang{R}

### 2004-11-16 paper/trunk/systemfit.tex
some minor changes and some comments on nlsystemfit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/systemfit@568 591db2bd-e0a0-4d22-8c19-2192dbb7da31

%% an abstract and keywords
\Abstract{ Many statistical analysis are based on models that consist
  of  more than one equation.  These equations should be fitted
  simultaneously rather than equation  by equation.  The package
  \pkg{systemfit} provides the capability to estimate systems  of
  linear and non-linear equations in \proglang{R}.  } \Keywords{R,
  simultaneous equation system, seemingly unrelated regression}

@@

This paper is organized as follows:  In section \ref{sec:Theory} we
introduce the mathematics of estimating equation systems and give a
comprehensive overview over all relevant issues regarding this topic.
Section \ref{sec:UsingSystemfit} demonstrates how to run
\pkg{systemfit}, especially how the features presented in the previous
chapter can be used.  Finally, a summary and an outlook are given in
section \ref{sec:Summmary}.

@@

@@

\begin{equation}
  \label{eq:non_linear_eq_1}
  \epsilon_{t} = q( y_t, x_t, \theta )
\end{equation}

@@

where, $r$ is a column vector for the residuals for each equation, $X$
is matrix of the partial derivatives with respect to the parameters,
$W$ is a matrix of the instrument variables, $Z(Z'Z)^{-1}Z$, $Z$ is a
matrix of the instrument variables, and $I$ is an $n \times n $
identity matrix and $S$ is the variance-covariance matrix between the
equations

@@

@@

%% Arne -- Did you put the ~'s in by hand?

@@

\subsubsection{nlsystemfit issue 1}
% The user should be aware that the function is \bold{VERY} sensative to
% the starting values and the nlm function may not converge. The nlm
% function will be called with the \code{typsize} argument set the
% absolute values of the starting values for the OLS and 2SLS
% methods. For the SUR and 3SLS methods, the \code{typsize} argument is
% set to the absolute values of the resulting OLS and 2SLS parameter
% estimates from the nlm result structre. In addition, the starting
% values for the SUR and 3SLS methods are obtained from the OLS and 2SLS
% parameter estimates to shorten the number of iterations. The number of
% iterations reported in the summary are only those used in the last
% call to nlm, thus the number of iterations in the OLS portion of the
% SUR fit and the 2SLS portion of the 3SLS fit are not included.  }

@@

\section{CVS Revision Log}
\begin{verbatim}
$Log$
Revision 1.8  2004/11/16 08:55:53  henningsena
merged file and resolved conflict due to some minor changes I did while Jeff was adding nlsystemfit.

>>>>

%% an abstract and keywords
\Abstract{ Many statistical analysis are based on models that consist
  of  more than one equation.  These equations should be fitted
  simultaneously rather than equation  by equation.  The package
  \pkg{systemfit} provides the capability to estimate systems  of
  linear and non-linear equations in \proglang{R}.
}

\Keywords{R, simultaneous equation system, seemingly unrelated regression}

@@

This paper is organized as follows: In section~\ref{sec:Theory} we
introduce the mathematics of estimating equation systems and give a
comprehensive overview over all relevant issues regarding this topic.
Section~\ref{sec:UsingSystemfit} demonstrates how to run
\pkg{systemfit}, especially how the features presented in the previous
chapter can be used. Finally, a summary and an outlook are given in
section~\ref{sec:Summmary}.

@@

The first line loads the \pkg{systemfit} package. 
The second line loads example data that are included in this package.
These data come from \cite{kmenta86}.
They are attached to the \proglang{R} search path in line three.
In the last line a seemingly unrelated regression is done.
The first equation represents the demand side of the food market.
The dependant variable is \code{q} (food consumption per capita). 
The regressors are \code{p} (ratio of food prices to general consumer prices)
and \code{d} (disposable income) as well as a constant%
\footnote{a regression constant is always implied if not explicitly omitted.}.
The second equation represents the supply side.
Variable \code{q} (food consumption per capita) is also the dependant 
variable of this equation. 
The regressors are again \code{p} (ratio of food prices to general 
consumer prices) and a constant as well as 
\code{f} (ratio of preceding year's prices received by farmers) and 
\code{a} (a time trend in years).
The regression result is assigned to the variable \code{fitsur}.

@@

\begin{equation}
  \label{eq:non_linear_eq_1}
  \epsilon_{t} = q( y_t, x_t, \theta )
\end{equation}
%%% It is not clear what these variable are. 
%%% Adding an equation representing only a single equation
%%% would make things clearer.

@@

where, $r$ is a column vector for the residuals for each equation,
%% what is the difference to \epsilon?
$X$ is matrix of the partial derivatives of the dependent variable 
with respect to the parameters $\left( \frac{ \partial y }{ \theta} \right)$,
%% is this correct?
%% in the linear section X is used for the regressors, 
%% thus we should use a different variable name here.
$W$ is a matrix of the instrument variables, $Z(Z'Z)^{-1}Z$, $Z$ is a
matrix of the instrument variables, and $I$ is an $n \times n $
identity matrix and $S$ is the estimated variance-covariance matrix between the
equations

@@

\textbf{The residual variance-covariance matrix can be calculated in 
different ways (section~\ref{sec:residcov}). 
It should be relatively easy to implement this also in nlsystemfit(), 
e.g. the function 'calcRCov' in systemfit() could be moved outside
systemfit() that it can be used also by nlsystemfit().}

@@

%% Arne -- Did you put the ~'s in by hand?
%% Yes, I did. I used find and replace to substitute the "~" for " ". 
%% This is necessary to retain the format of the output.

@@

\subsubsection{nlsystemfit issue 1}
% The user should be aware that the function is \bold{VERY} sensative to
% the starting values and the nlm function may not converge.

% did you try optim()? Especially if you provide (analytical) derivatives
% it is very good in finding the optimum.

% The nlm
% function will be called with the \code{typsize} argument set the
% absolute values of the starting values for the OLS and 2SLS
% methods. For the SUR and 3SLS methods, the \code{typsize} argument is
% set to the absolute values of the resulting OLS and 2SLS parameter
% estimates from the nlm result structre. In addition, the starting
% values for the SUR and 3SLS methods are obtained from the OLS and 2SLS
% parameter estimates to shorten the number of iterations. The number of
% iterations reported in the summary are only those used in the last
% call to nlm, thus the number of iterations in the OLS portion of the
% SUR fit and the 2SLS portion of the 3SLS fit are not included.  }

@@

\section{CVS Revision Log}
\begin{verbatim}
$Log$
Revision 1.9  2004/11/16 10:59:09  henningsena
some minor changes and some comments on nlsystemfit

Revision 1.8  2004/11/16 08:55:53  henningsena
merged file and resolved conflict due to some minor changes I did while Jeff was adding nlsystemfit.
